Chain separation distribution in protein contact networks

### 2021-12-03 paper/main.tex
fixing folder

\begin{document}
\title{Chain separation distribution in protein contact networks}
\title{Theoretical approximation of chain separation distribution in protein contact networks}
\title{Theory of chain separation distributions gives insights in real protein contact networks}
%\title{}
\author{Nora Molkenthin}
\affiliation{Potsdam Institut für Klimafolgenforschung}
\affiliation{Center for Advancing Electronics Dresden (cfaed), Technical University of Dresden}
\author{J. J.  Güven}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}
\author{Steffen Mühle}
\affiliation{Uni Göttingen, 37077 Göttingen, Germany}
\author{Antonia S J S Mey}
\email[Electronic Address: ]{antonia.mey@ed.ac.uk}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}

@@

Here, we show how a simple geometrical model can reproduce this heuristically found behaviour well, and even provide a simple analytical approximation. The three-dimensional geometrical model and the analytical approximation of its two-dimensional analogue both build on the idea that inherently geometrical objects, such as amino acids, imply that any resulting PCN network ensemble modelling them has to be spatially embedded~\cite{molkenthin2016scaling, molkenthin2020self}. 
Different approaches have been used in the past for geometrical models describing protein folding, some of which derived characteristics of the secondary structure from constraints on bond and torsion angles~\cite{bhattacharjee2013flory,Danielsson2010,Molkenthin2011} and others focused on the formation mechanisms of the tertiary structure~\cite{molkenthin2016scaling, molkenthin2020self}.

@@

The underlying formation mechanisms of individual protein folds are incredibly complex and depend on the surrounding solvent as well as the specific amino acid sequence and their quantum-mechanical interactions. Here we want to study the ensemble of all folded proteins, however, and will thus "average" over such specifics.

Based on the basic observation that amino acids are objects in space which can not overlap indefinitely, we introduced a model in \cite{molkenthin2020self} starting from a chain of identical spheres, each in contact only with the neighbours it is connected to. Then additional connections form by randomly selecting two spheres and moving them towards each other until they touch, the new links formed that way can not be broken in later steps and function as constraints on subsequent links. If contact of the two selected spheres is geometrically impossible without breaking previously made connections or leading to overlaps of any spheres, the link is not made and taken out of the pool of possible connections. This process is repeated until no more links can be formed without violating the geometric constraints.

In \cite{molkenthin2016scaling}, we have introduced a simplified, two-dimensional version of this model, which starts from a closed chain (i.e. a ring) of $N$ unit discs and subsequently adds links, such that connected discs touch, yet no discs intersect. The advantage of this model is that it can be approximated by a purely topological simplification, which can be treated analytically.

@@

The expected distribution of possible links after one step is given thus by the average over available link pools taken over all possible lengths of the initial link
\begin{align}
    F_1(s)&=\frac{1}{N/2-1} \sum_{s_1=2}^{N/2}{ \begin{cases}
    F_0(s)-2(s-1) \text{ , for } s<s_1\\
    F_0(s)-2(s_1 -1)\text{ , for } s\geq s_1
    \end{cases}}
    \nonumber \\
    &\approx\frac{2}{N} \sum_{s_1=2}^{N/2} { \begin{cases}
    N-2(s-1) \text{ , for } s<s_1\\
    N-2(s_1 -1)\text{ , for } s\geq s_1
    \end{cases}}
\end{align}

In subsequent steps we take the weighted average over new links being added at step $k-1$, where the probability that a link of length $s$ can still be made is
\begin{equation}
    P_{in}^k(s)=F_k(s)/C_k,
\end{equation}
where $C_k=\sum_{s=2}^{N/2}F_k(s)$ and get the same reduction (compare Fig.~\ref{fig:schematic}) but starting from the pool $F_{k-1}(s)$ of the step before rather than the full $N$ links for each length. To account for possible overlaps between newly removed links and links that were removed previously, we multiply by
\begin{equation}
    \frac{F_k(s)}{F_0(s)} \approx F_k(s)\frac{1}{N}.
    \label{eq.Pk}
\end{equation}
% \begin{equation}
%     P_k(s)=\frac{F_k(s)}{C_k},
%     \label{eq.Pk}
% \end{equation}
%with $C_k=\sum_{s=2}^{N/2}F_k(s)$, 
This ignores any bias in which links are removed, however, in small, manageable examples this assumption seemed to hold well. 
This leads to the recursion
\begin{align}
    F_k(s)&= \sum_{s_k=2}^{N/2} P_{in}^k(s_k)\nonumber \\
    & (F_{k-1}(s)-  F_k(s)\frac{1}{N}
    {\begin{cases}
     2(s-1) \text{ , for } s<s_k\\
     2(s_k -1))\text{ , for } s\geq s_k
    \end{cases}} )
\end{align}
which simplifies to
\begin{align}
   F_k(s)&=F_{k-1}(s)(1-\frac{1}{N} \sum_{s_k=2}^{s} (2 (s_k-1) P_{in}^k(s_k))\nonumber \\
   &-\frac{1}{N} \sum_{s_k=s+1}^{N/2} (2 (s-1) P_{in}^k(s_k)))\nonumber \\
   &=F_{k-1}(s)(1-\frac{1}{N}
   \sum_{s_k=2}^{s} (2 (s_k-1) P_{in}^k(s_k))\nonumber \\
   &-\frac{2 (s-1)}{N} (\sum_{s_k=2}^{N/2} P_{in}^k(s_k) - \sum_{s_k=2}^{s} P_{in}^k(s_k)))\nonumber \\
   &=F_{k-1}(s)(1-\frac{2}{N} (s-1) 
   +\frac{2}{N} s
   \sum_{s_k=2}^{s}P_{in}^k(s_k)\nonumber \\
   &-\frac{2}{N}
   \sum_{s_k=2}^{s} s_k P_{in}^k(s_k))
   \label{eq.Fk_rec}
\end{align}

To approximately evaluate this we make a guess for $P_{in}^k(s)$ at low $k$ and at high $k$, namely
\begin{equation}
    P_{in}^{k\ll N/2}(s)\approx P_{in}^{k=0}(s)=2/N.
    \label{eq.P_small}
\end{equation}
For later steps $k$, the probability of a link to still be in the pool decreases with $s$, we thus make the ansatz, that it drops off as $1/s$, leading to
\begin{equation}
    P_{in}^{k\gg 1}(s)\approx \frac{s^{-1}}{ (H_{\frac{N}{2}}-1)},
    \label{eq.P_large}
\end{equation}
where $H_{\frac{N}{2}}$ is the harmonic number. 
We suspect that the approximation for low $k$ only holds for a couple of steps while Eq.~\ref{eq.P_large} holds, at least on average for larger $k$. We have no approximation for the intermediate steps but assume that the errors roughly cancel each other out. Inserting Eq.\ref{eq.P_small} and Eq.\ref{eq.P_large} into Eq.\ref{eq.Fk_rec} results in the following two recursions for the early and late evolution of the sequence distance distribution respectively.
\begin{align}
   F_{k\ll N/2}(s)&\approx 
   F_{k-1}(s)(1-\frac{2}{N} (s-1) 
   +\frac{4}{N^2} s (s-1)\nonumber \\
   &-\frac{2}{N^2}(s^2+s-2)\nonumber \\
   &=F_{k-1}(s)(1-f_{low}(s)),
   \label{eq.Fk_rec_low}
\end{align}
where 
\[f_{low}(s)=-\frac{2}{N^2}s^2+(\frac{2}{N}+\frac{6}{N^2})s-(\frac{2}{N}+\frac{4}{N^2}),\]
and
\begin{align}
   F_{k\gg1}(s)&\approx 
   F_{k-1}(s)(1-\frac{2}{N} (s-1) 
   +\frac{2}{N} s \sum_{s_k=2}^{s}\frac{1}{s_k(H_{\frac{N}{2}}-1)}\nonumber \\
   &-\frac{2}{N}
   \sum_{s_k=2}^{s} s_k \frac{1}{s_k(H_{\frac{N}{2}}-1)})\nonumber \\
   &=
   F_{k-1}(s)(1 -(\frac{2}{N}-\frac{2}{N}\frac{H_{s}-1}{H_{\frac{N}{2}}-1}+\frac{2}{N(H_{\frac{N}{2}}-1)})s \nonumber \\ 
   &+ \frac{2}{N(H_{\frac{N}{2}}-1)} + \frac{2}{N})\nonumber \\
   &=F_{k-1}(s)(1-f_{high}(s)).
   \label{eq.Fk_rec_high}
\end{align}
where 
\[f_{high}(s)=\frac{2}{N}(\frac{H_{\frac{N}{2}}-H_{s}+1}{H_{\frac{N}{2}}-1}s-\frac{H_{\frac{N}{2}}}{H_{\frac{N}{2}}-1}).\]

We can now use the recursive expressions to write down a closed expression for $F_k(s)$

\begin{align}
    F_k(s)&=N\prod_{i_1}^a\left(1-f_{low}(s)\right)\prod_{i=a+1}^k\left(1-f_{high}(s)\right) \nonumber \\
    &= N\left(\frac{1-f_{low}(s)}{1-f_{high}(s)}\right)^a\left(1-f_{high}(s)\right)^{k}
\end{align}

In 2D this is repeated for $N-3$ steps until no more links can be added \cite{molkenthin2016scaling}.
The probability distribution of the realized residue distances of all added links is then given by the average over the available pools at each link addition step, leading to
\begin{align}
    P(s)&=\frac{1}{N-3}\sum_{k=0}^{N-4} \frac{F_k(s)}{C_k} \nonumber \\
    &= \frac{N}{N-3}\left(\frac{1-f_{low}(s)}{1-f_{high}(s)}\right)^a
\sum_{k=0}^{N-4} \frac{1}{C_k} \left(1-f_{high}(s)\right)^{k} \nonumber \\
    &\approx \left(\frac{1-f_{low}(s)}{1-f_{high}(s)}\right)^a\frac{A}{f_{high}(s)},
    \label{eq.Ps}
\end{align}

\red{The assumption that the sum is basically a harmonic series seems like a bit of a stretch I shall check it again but the results look good}

for $1<s<\frac{N}{2}$. The approximation sets all $C_k=C$ to allow the use of the geometric series for solving the equation. We can do this because $C_k$ monotonically decreases with $k$ and $N>C_k>C_{N-2}$ meaning we can obtain an upper and lower bound by replacing all $C_k$ in the expression with either $N$ (underestimates all probabilities) or $C_{N-2}$ (overestimates all probabilities), however, as neither of those depend on $s$, these bounds are merely in the constant factor $A$ and do not affect the scaling as such.
%which makes use of Eq.~\ref{eq.Pk}. 
%The only unknown in this expression is $C_k$, \red{which we can not compute exactly...?}. However, since $C_k$ monotonically decreases with $k$, $N>C_k>C_{N-2}$ always holds and we can obtain an upper and lower bound by replacing all $C_k$ in the expression with either $N$ (underestimates all probabilities) or $C_{N-2}$ (overestimates all probabilities). Each time we get an expression of the form 
% \begin{align}
%      P(s)&\approx\frac{N}{N-3}\sum_{k=0}^{N-3} \frac{1}{C_k}\prod_{i}^k\left(1-\frac{f(s)}{(\frac{N}{2}-2)C} \right)\nonumber \\
%      &=\frac{N}{N-3} \sum_{k=0}^{N-3}\frac{1}{C_k}\left(1-\frac{f(s)}{(\frac{N}{2}-2)C} \right)^k \nonumber \\
%      &\approx \frac{A}{f(s)}
% \end{align}
% for $1<s<\frac{N}{2}$,
% where $A$ is an unknown constant, containing an approximation of the various normalization constants $C_i$, as well as the direct effects of the chain length. 
Since for long chains and $s \ll N/2$, $f(s)$ is dominated by the linear term, the distribution is well approximated by a power law with an exponent of $-1$.

%In the last step of this approximation, we took the $C_k$ to all be approximately equal in order to apply the geometric series. However, in reality we know $1/C_k$ to increase with $k$, giving large $k$ summands relatively more weight than in a geometric series. This can be expressed as a small correction $\alpha$ in the exponent.

%\begin{equation}
 %   P_{corrected}(s)\approx A f(s)^{-1+\alpha},
%\end{equation}
%where $0<\alpha \ll 1$. \blue{The parameters $A$ and $\alpha$ in this approximation are treated as free parameters and used for fitting the theoretical result to data and simulations. While they could be estimated by evaluating the middle line of Eq.~\ref{eq.Ps}, this estimate is for the 2D model, in which the exclusion of connections is faster, leading to a steeper decrease of $C_k$, as well as a different factor $A$.}

@@

\section*{Data and simulations}
For large values of the chain length $N$ we thus find an approximate power law probability distribution for the residue distance with an exponent of $-1+\alpha$. Here we use simulations, as introduced in~\cite{molkenthin2020self}, and networks extracted from measured protein data from the PDB \cite{PDB} to compare this behaviour with the more realistic 3D settings and show that the basic principle persists, indicating that much of the residue distance distribution can be explained by generic geometric constraints.
 \begin{figure}[h]
        \centering
	\includegraphics[width=0.45\textwidth]{paper/figures/Fig2/pdb_statistics.pdf}
        \caption{Statistics on all PDB entries based on the number of amino acid residues, with most entries found in the interval of 100-400 residues as of May 26th 2020.}
        \label{fig:pdb_stats}
\end{figure}

Protein chain lengths in the PDB vary from small fragments of less than 10 amino acids to large agglomerates of over 2000 residues. The most frequently occurring chain lengths, however, are between 100 and 400 residues as shown in Fig.~\ref{fig:pdb_stats}. \red{For our analysis, we have thus chosen ............ structures of around 200, ........ structures of around 300 and .......... structures of around 400 residues from the PDB.} Those have been converted into PCN's based on the measured locations of their residues to analyze the distribution of residue distances between connected amino acids. The resulting distribution is compared to those generated in simulations in~\cite{molkenthin2020self} as well as the above derived theoretical expression scaled with a factor to account for the different dimensionality, as well as approximations made in the derivation.

>>>>

\begin{document}
\title{Chain separation distribution in protein contact networks}
\title{Theoretical approximation of chain separation distribution in protein contact networks}
\title{Theory of chain separation distributions gives insights in real protein contact networks}
%\title{}
\author{Nora Molkenthin}
\affiliation{Potsdam Institut für Klimafolgenforschung}
\affiliation{Center for Advancing Electronics Dresden (cfaed), Technical University of Dresden}
\author{Steffen Mühle}
\affiliation{Uni Göttingen, 37077 Göttingen, Germany}
\author{Antonia S J S Mey}
\email[Electronic Address: ]{antonia.mey@ed.ac.uk}
\affiliation{EaStCHEM School of Chemistry, University of Edinburgh, Edinburgh, United Kingdom}
%\author{Marc Timme}
% \affiliation{Network Dynamics, Max Planck Institute for Dynamics and Self-Organization (MPIDS), 37077 Göttingen, Germany}
% \affiliation{Chair for Network Dynamics, Institute for Theoretical Physics and Center for Advancing Electronics Dresden (cfaed), Technical University of Dresden, 01069 Dresden}

@@

Here, we show how a simple geometrical model can reproduce this heuristically found behaviour well, and even provide a simple analytical approximation. \blue{The three-dimensional geometrical model and the analytical approximation of its two-dimensional analogue both} build on the idea that inherently geometrical objects, such as amino acids, imply that any resulting PCN network ensemble modelling them has to be spatially embedded~\cite{molkenthin2016scaling, molkenthin2020self}. 
Different approaches have been used in the past for geometrical models describing protein folding, some of which derived characteristics of the secondary structure from constraints on bond and torsion angles~\cite{bhattacharjee2013flory,Danielsson2010,Molkenthin2011} and others focused on the formation mechanisms of the tertiary structure~\cite{molkenthin2016scaling, molkenthin2020self}.

@@

The underlying formation mechanisms of individual protein folds are incredibly complex and depend on the surrounding solvent as well as the specific amino acid sequence and their quantum-mechanical interactions. Here we want to study the ensemble of all \blue{folded(?)} proteins, however, and will thus "average" over such specifics.

Based on the basic \blue{observation} that amino acids are objects in space which can not overlap indefinitely, we introduced a model in \cite{molkenthin2020self} starting from a chain of identical spheres, each in contact only with the neighbours it is connected to. Then additional connections form by randomly selecting two spheres and moving them towards each other until they touch, the new links formed that way can not be broken in later steps and function as constraints on subsequent links. If contact of the two selected spheres is geometrically impossible without breaking previously made connections or leading to overlaps of any spheres, the link is not made and taken out of the pool of possible connections. This process is repeated until no more links can be formed without violating the geometric constraints.

In \cite{molkenthin2016scaling}, we have introduced a simplified, two-dimensional version of this model, which starts from a closed chain \blue{(i.e. a ring)} of $N$ unit discs and subsequently adds links, such that connected discs touch, yet no discs intersect. The advantage of this model is that it can be approximated by a purely topological simplification, which can be treated analytically.

@@

The expected distribution of possible links after one step is given thus by the average over available link pools taken over all possible lengths of the initial link
\blue{\begin{align}
    F_1(s)&=\frac{1}{N/2-1} \sum_{s_1=2}^{N/2}{ \begin{cases}
    F_0(s)-2(s-1) \text{ , for } s<s_1\\
    F_0(s)-2(s_1 -1)\text{ , for } s\geq s_1
    \end{cases}}
    \nonumber \\
    &\approx\frac{2}{N} \sum_{s_1=2}^{N/2} { \begin{cases}
    N-2(s-1) \text{ , for } s<s_1\\
    N-2(s_1 -1)\text{ , for } s\geq s_1
    \end{cases}}
\end{align}}
Here we denote the size of the pool of available links after $i$ links have been added $F_i(s)$.

In subsequent steps we \blue{still average over $N/2-1$ possible link lengths made at step $k-1$, each occurring with probability $F_k(s)/C_k$, where $C_k=\sum_{s=2}^{N/2}F_k(s)$ and} get the same reduction \blue{(compare Fig.~\ref{fig:schematic})} but starting from the pool $F_{k-1}(s)$ of the step before rather than the full $N$ links for each length. The reduction also needs to be multiplied by a factor \blue{accounting for already removed links}
\blue{\begin{equation}
    P_k(s)=\frac{F_k(s)}{N/2-1} \approx F_k(s)\frac{2}{N},
    \label{eq.Pk}
\end{equation}}
% \begin{equation}
%     P_k(s)=\frac{F_k(s)}{C_k},
%     \label{eq.Pk}
% \end{equation}
%with $C_k=\sum_{s=2}^{N/2}F_k(s)$, 
to only remove links that are still in the pool. This leads to the recursion
\blue{\begin{align}
    F_k(s)&= \sum_{s_k=2}^{N/2} F_k(s)/C_k\nonumber \\
    &{\begin{cases}
     F_{k-1}(s)-2(s-1) P_{k-1}(s) \text{ , for } s<s_k\\
     F_{k-1}(s)-2(s_k -1)P_{k-1}(s)\text{ , for } s\geq s_k
    \end{cases}}\nonumber \\
    &\approx \frac{1}{N/2} \sum_{s_k=2}^{N/2} \nonumber \\
    &{\begin{cases}
     F_{k-1}(s)-2(s-1) P_{k-1}(s) \text{ , for } s<s_k\\
     F_{k-1}(s)-2(s_k -1)P_{k-1}(s)\text{ , for } s\geq s_k
    \end{cases}},
    \label{eq.Fk_def}
\end{align}}
\blue{where we made the approximation that the probabilities $F_k(s)/C_k\approx 2/N$ remain approximately uniform. This approximation holds best for small $s$ and large $N$.} \red{Toni and Steffen, does either of you have access to a Mathematica license? I could start bugging my admin about getting one but i think that would probably take longer than we want to spend on this. If we find a license we could probably take the first line of eq. 4 and get a plot out using Rsolve or something...  And plot the result with the approximation.}

\blue{We further simplify Eq.~\ref{eq.Fk_def}.}

\blue{\begin{align}
   F_k(s)&\approx \frac{4}{N^2} F_{k-1}\nonumber \\
   &\left( \sum_{s_k=2}^{N/2}\frac{N}{2} - \sum_{s_k=2}^{s} 2(s_k-1) - \sum_{s_k=s+1}^{N/2} 2(s -1) \right) \nonumber \\
   &= \frac{4}{N^2} F_{k-1}\nonumber \\
   &\left(\frac{N^2}{4} -(s^2-s)
   -2\left(\frac{N}{2}-(s+1)\right)(s-1)\right)  \nonumber \\
   &= F_{k-1}(s)\left(1-\frac{-s^2 +(N-1)s-(N+2)}{\frac{N^2}{4}} \right)\nonumber \\
   &=F_{k-1}(s)\left(1-\frac{f(s)}{\frac{N^2}{4}} \right)
   \label{eq.Fk_rec}
\end{align}}
% \begin{align}
%   F_k(s)&= \frac{1}{\frac{N}{2}-2} \frac{F_{k-1}(s)}{C_{k-1}}\nonumber \\
%   &\left( \sum_{s_k=2}^{N/2}C_{k-1} - \sum_{s_k=2}^{s} 2(s_k-1) - \sum_{s_k=s+1}^{N/2} 2(s -1) \right) \nonumber \\
%   &= \frac{1}{\frac{N}{2}-2}\frac{F_{k-1}(s)}{C_{k-1}}\nonumber \\
%   &\left(\left(\frac{N}{2}-2\right)C_{k-1} -(s^2-s)
%   -2\left(\frac{N}{2}-(s+1)\right)(s-1)\right)  \nonumber \\
%   &= F_{k-1}(s)\left(1-\frac{-s^2 +(N-1)s-(N+1)}{(\frac{N}{2}-2)C_{k-1}} \right)\nonumber \\
%   &=F_{k-1}(s)\left(1-\frac{f(s)}{(\frac{N}{2}-2)C_{k-1}} \right)
%   \label{eq.Fk_rec}
% \end{align}
Where we defined $f(s)=-s^2 +(N-1)s-(N+2) \approx N s - s^2 - N$ for notational brevity.

We can now use the recursive expression Eq.~\ref{eq.Fk_rec} to write down a closed expression for $F_k(s)$

\blue{\begin{equation}
    F_k(s)=N\prod_{i}^k\left(1-\frac{f(s)}{\frac{N^2}{4}} \right) = N \left(1-\frac{f(s)}{\frac{N^2}{4}} \right)^k
\end{equation}}
% \begin{equation}
%     F_k(s)=N\prod_{i}^k\left(1-\frac{f(s)}{(\frac{N}{2}-2)C_{i-1}} \right)
% \end{equation}

\blue{In 2D this is repeated for $N-3$ steps until no more links can be added \cite{molkenthin2016scaling}.}
The probability distribution of the realized residue distances of all added links is then given by the average over the available pools at each link addition step, leading to
\blue{\begin{align}
    P(s)&=\frac{1}{N-3}\sum_{k=0}^{N-4} \frac{F_k(s)}{C_k} \nonumber \\
    &= \frac{N}{N-3}\sum_{k=0}^{N-4} \frac{1}{C_k} \left(1-\frac{f(s)}{\frac{N^2}{4}} \right)^k \nonumber \\
    &\approx \frac{A}{f(s)},
    \label{eq.Ps}
\end{align}}
% \begin{align}
%     P(s)&=\frac{1}{N-3}\sum_{k=0}^{N-3} P_k(s) \nonumber \\
%     &= \frac{N}{N-3}\sum_{k=0}^{N-3} \frac{1}{C_k}\prod_{i}^k\left(1-\frac{f(s)}{(\frac{N}{2}-2)C_{i-1}} \right),
% \end{align}
\blue{for $1<s<\frac{N}{2}$. The approximation sets all $C_k=C$ to allow the use of the geometric series for solving the equation. We can do this because $C_k$ monotonically decreases with $k$ and $N>C_k>C_{N-2}$ meaning we can obtain an upper and lower bound by replacing all $C_k$ in the expression with either $N$ (underestimates all probabilities) or $C_{N-2}$ (overestimates all probabilities), however, as neither of those depend on $s$, these bounds are merely in the constant factor $A$ and do not affect the scaling as such.}
%which makes use of Eq.~\ref{eq.Pk}. 
%The only unknown in this expression is $C_k$, \red{which we can not compute exactly...?}. However, since $C_k$ monotonically decreases with $k$, $N>C_k>C_{N-2}$ always holds and we can obtain an upper and lower bound by replacing all $C_k$ in the expression with either $N$ (underestimates all probabilities) or $C_{N-2}$ (overestimates all probabilities). Each time we get an expression of the form 
% \begin{align}
%      P(s)&\approx\frac{N}{N-3}\sum_{k=0}^{N-3} \frac{1}{C_k}\prod_{i}^k\left(1-\frac{f(s)}{(\frac{N}{2}-2)C} \right)\nonumber \\
%      &=\frac{N}{N-3} \sum_{k=0}^{N-3}\frac{1}{C_k}\left(1-\frac{f(s)}{(\frac{N}{2}-2)C} \right)^k \nonumber \\
%      &\approx \frac{A}{f(s)}
% \end{align}
% for $1<s<\frac{N}{2}$,
% where $A$ is an unknown constant, containing an approximation of the various normalization constants $C_i$, as well as the direct effects of the chain length. 
\blue{Since for long chains and $s \ll N/2$, $f(s)$ is dominated by the linear term, the distribution is well approximated by a power law with an exponent of $-1$.}

In the last step of this approximation, we took the $C_k$ to all be approximately equal in order to apply the geometric series. However, in reality we know $1/C_k$ to increase with $k$, giving large $k$ summands relatively more weight than in a geometric series. This can be expressed as a small correction $\alpha$ in the exponent.

\begin{equation}
    P_{corrected}(s)\approx A f(s)^{-1+\alpha},
\end{equation}
where $0<\alpha \ll 1$. \blue{The parameters $A$ and $\alpha$ in this approximation are treated as free parameters and used for fitting the theoretical result to data and simulations. While they could be estimated by evaluating the middle line of Eq.~\ref{eq.Ps}, this estimate is for the 2D model, in which the exclusion of connections is faster, leading to a steeper decrease of $C_k$, as well as a different factor $A$.}

@@

\section*{Data and simulations}
For large values of the chain length $N$ we thus find an approximate power law probability distribution for the residue distance with an exponent of $-1+\alpha$. Here we use simulations, as introduced in~\cite{molkenthin2020self}, and networks extracted from measured protein data from the PDB \cite{PDB} to compare this behaviour \blue{with the more} realistic 3D settings and show that the basic principle persists, indicating that much of the residue distance distribution can be explained by simple\blue{/generic?} geometric constraints.
 \begin{figure}[h]
        \centering
	\includegraphics[width=0.45\textwidth]{paper/figures/Fig2/pdb_statistics.pdf}
        \caption{Statistics on all PDB entries based on the number of amino acid residues, with most entries found in the interval of 100-400 residues as of May 26th 2020.}
        \label{fig:pdb_stats}
\end{figure}

Protein chain lengths in the PDB vary from small fragments of less than 10 amino acids to large agglomerates of over 2000 residues. The most frequently occurring chain lengths, however, are between 100 and 400 residues as shown in Fig.~\ref{fig:pdb_stats}. \red{For our analysis, we have thus chosen ............ structures of around 200, ........ structures of around 300 and .......... structures of around 400 residues from the PDB.} Those have been converted into PCN's based on the measured locations of their residues to analyze the distribution of residue distances between connected amino acids. The resulting distribution is compared to those generated in simulations in~\cite{molkenthin2020self} \blue{as well as the above derived theoretical expression scaled with a factor to account for the different dimensionality, as well as approximations made in the derivation.}
